### 2019-12-09 ch10_discussion/content.tex
~
updates
work on discussion

In this work, we aimed to accelerate scientific progress in the field of single-cell omics by providing and evaluating computational tools. In section~\ref{sec:research_objectives} we proposed a concrete set of objectives this work will attempt to solve. Every chapter respectively tackles one of these research objectives, and each chapter discusses the implications of that part of our research in detail. This chapter reflects on the impact of this work on the field, going over each of the objectives and respective chapters.

\section{Trajectory inference}

\section{Network inference}

\section{Benchmarking in bioinformatics}

\section{Benchmarking with \textit{in silico} single cells}

We developed \texttt{dyngen}, a simulator of single cells to benchmark and stress-test computational tools for single-cell omics (Chapter~\ref{chap:dyngen}).
The \texttt{dyngen} software supports a selection of existing experimental protocols, and can also simulate experiments that are not possible with the current technologies, such as generating omics profiles from the same cell at multiple timepoints.
The generated data can help kick-start emerging domains with low data availability more safely by allowing software developers to test their method before numerous datasets become publicly available.

\texttt{dyngen} has already been successfully used to evaluate trajectory inference \cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment \cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference \cite{pratapa_benchmarkingalgorithmsgene_2019} methods. Furthermore, we showed that it can also be used to evaluate differential network inference methods, and trajectory alignment methods.

% Eventueel kort ook beschrijven waar je denkt dat het nog van toepassing kan zijn? The use if dyngene is not limited to.... but might also,....
% Eventueel ook: waar zie je nog verbeter/uitbreidingspunten in je methode. 
% Wat zijn de huidige limitaties. Bv. Is er een tijdgebonden limitatie. bv. Stel je wilt data van 10.000 cellen genereren maar bent een maand bezig, etc.

\section{Comparing 45 trajectory inference methods}
Using this simulator and a collection of real datasets, we performed a comparison of 45 TI methods (Chapter~\ref{chap:dynbenchmark}).
Our contributions include writing software to run 45 different error-prone TI methods with a common interface, downloading and processing hundreds of single cell datasets, and developing novel metrics for comparing ground truth and predicted trajectories.
With this benchmarking study, we had two clear goals in mind: to help guide users to TI methods that are suitable for their needs, and to provide developers of TI methods with the necessary tools to benchmark their own tool.

We accomplished the first goal by constructing a set of guidelines for end-users to help choose a TI method that is accurate, robust and fit for their application. 
Since such guidelines were hitherto lacking, they are now commonly disseminated in manuscripts \cite{lafzi_tutorialguidelinesexperimental_2018,luecken_currentbestpractices_2019}, courses \cite{kiselev_analysissinglecell_2019,martens_analysissinglecell_2019}, and slides shown during keynote caffeine refuelling sessions \cite{hemberg_coffeebreakanalysis_2019}.

For our second goal, it seems too early to tell whether our research had any effect on self-assessments of TI method developers. We did made our pipeline, datasets, metrics, and containerised wrappers of TI methods publicly available for developers to use.
However, so far we do not observe an increase in developers performing quantitative benchmarks. We hypothesise causal reasons for this phenomenon and provide solutions in order to spur TI developers to perform more self-assessments (Chapter~\ref{chap:selfassessment}).
% "We hypothesise." Where/what do you hypothise? In your paper? Dan zou ik verwijzen naar deze sectie.
% Eventueel ook: Is there room for improvement for existing methods (deze zullen vermoedelijk onder een andere naam gepubliceerd dienen te worden)
% Dienen eerdere publicaties in vraag te worden gesteld (indien minder goede methods gebruikt werden?
% Kan dit uitgebreid worden naar andere methoden?
% Is dyngen niet ontstaan omdat je nood had aan betere datasets? Zou je in dit opzicht niet dit deeltje eerst plaatsen en verwijzen naar dyngen?
% Misschien ook vermelden dat de publicatie een aanzienlijke support verwacht van jullie team na publicatie. Dat de hoge nood gepaard gaat met een grote vraag. 
\section{A toolkit to infer, visualise and interpret single-cell trajectories}

The previous two works necessitated developing a tool for visualising trajectory data. We extended this implementation into a full toolkit, named \texttt{dyno}, for inferring and analysing trajectories (Chapter~\ref{chap:dyno}). The toolkit allows performing downstream analyses such as detecting differentially expressed genes along transitions in the trajectory, annotating the trajectory, and comparing multiple trajectories in a common dimensionality reduction. A major benefit of \texttt{dyno} is that it allows utilising any of the 50 TI methods we wrapped as part of the comparison, without having to install dependencies for every TI method separately.

The software packages consist of several subpackages, each with a separate functionality (e.g. \texttt{dynplot} visualises trajectories, \texttt{dynmethods} provides wrappers for the TI methods). Since most of the packages are currently only hosted on Github, only users familiar with Github have been able to use it. To make \texttt{dyno} available to a larger audience, we are publishing each of these packages and developing command-line interfaces for the most important functionalities.

% Are we talking about R-packages?

\section{Fast, accurate, and robust single-cell pseudotime}

We developed a TI method, called \texttt{SCORPIUS}, specialised in inferring linear trajectories (Chapter~\ref{chap:scorpius}). Since our comparison of TI methods already lists 30 TI methods that can infer trajectories that are more complex than a linear ordering of cells, why would we need a linear TI method?

Being able to study how cells progress over time is in terms of analysis as fundamental as being able to cluster cells or to identify differentially expressed genes between two groups. We showed that for analysing datasets containing linear trajectories, \texttt{SCORPIUS} outperforms all other TI methods tested as part of our TI benchmark.

Despite \texttt{SCORPIUS} published on bioRxiv in 2016, discussion and usage of the pre-print has largely been limited to comparing the results of a different (linear) TI methods to those of \texttt{SCORPIUS}. Since the pre-print, many improvements have been made to improve the accuracy and robustness of predicted trajectories, to allow scaling up to datasets containing 100'000s of cells, and to providing a more user-friendly interface. Since the beginning of 2019 -- before the publication of our benchmark of TI methods -- several studies have now reported using \texttt{SCORPIUS} as a tool for deriving primary results \cite{damond_maphumantype_2019, cheng_multifactorialheterogeneityvirusspecific_2019, tibbitt_singlecellrnasequencing_2019, hove_singlecellatlasmouse_2019, wouters_singlecellgeneregulatory_2019, reimegard_combinedmrnaprotein_2019}.

\section{Inferring single cell regulatory networks}

We proposed a novel algorithm for inferring case-wise gene regulatory interactions (Chapter~\ref{chap:bred}). While we demonstrated the methodology by inferring 14'963 sample-specific GRNs from the TCGA project, further interpretation of these results needs to be perform to validate the usefulness of the method.

Additional avenues for exploration could involve applying trajectory inference on the case-wise regulome data, and quantitatively evaluating the performance of the algorithm using the \texttt{dyngen} simulations.

\section{Optimising regulatory networks}
This project was a spin-off of Netter \cite{ruyssinck_netterrerankinggene_2016}, a tool for reranking GRN predictions using structural network properties. 
Predicted GRNs often are topologically dissimilar to real networks, and as such, Netter quantifies the topological properties of a network and makes incremental changes in order to arrive at a network with a 'better' topological profile.

The topological 'profile' of a network was defined as the frequencies of particular topological patterns called 'graphlets'. Keeping track of the frequencies of graphlets during an iterative process proved difficult at first. 
\texttt{incgraph} provides a solution to this problem (Chapter~\ref{chap:incgraph}) by  assuming that the incremental changes between two different networks are relatively small (less than 100 edge additions/removals). Under this assumption, it is possible to calculate the differences in topological patterns significantly faster in comparison to recomputing the graphlet frequencies from scratch at every iteration.

While \texttt{incgraph} solved a very specific and interesting problem, and was eventually published as a peer-reviewed article, it has not succeeded in making a direct impact on scientific research as of yet. Perhaps, at some point, a researcher will come across \texttt{incgraph} and find that it exactly solves his or her problem.

% Of je dient voor te stellen bij welke dataset het goed van toepassing is en het te promotor onder de Vlaamse research gemeenschap?

>>>>

\section{Impact of this work}
In this work, we made scientific contributions to the field of computational biology applied to single-cell omics. More specifically, we developed approaches for performing trajectory inference and network inference analyses for single-cell omics, as well as approaches for assessing the performance of such methods in a quantitative way.

The contribution with the largest impact in the field is the large-scale comparison of 45 TI methods. Based on our results, we provided guidelines on how to perform a TI analysis and developed a toolkit for performing a inferring and interpreting trajectories using any of these methods. Since such guidelines were hitherto lacking, they are now commonly disseminated in manuscripts \cite{lafzi_tutorialguidelinesexperimental_2018, luecken_currentbestpractices_2019}, courses \cite{kiselev_analysissinglecell_2019, martens_analysissinglecell_2019}, and slides shown during keynote caffeine refuelling sessions \cite{hemberg_coffeebreakanalysis_2019}. These disseminations are having a significant impact in how TI analyses are being performed in academia and industry alike.

This dissertation commences with a rant on low self-assessment rates of method developers. As a result, we developed a simulator of \textit{in silico} single cells, that allow quantifying the accuracy the prediction of a tool, even if the real data required to perform this analysis does not exist yet. While the manuscript is not published yet, dyngen has already been used to evaluate trajectory inference \cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory-based differential expression \cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference \cite{pratapa_benchmarkingalgorithmsgene_2019} methods.

% has used dyno detomaso_functionalinterpretationsingle_2019

\section{Outlook}
The types of computational analyses presented in Section~\ref{sec:comptools} (Clustering, DE, DR, TI, NI) all become routine methodologies for analysing and interpreting single-cell omics data. Here, we discuss several promising categories of computational methods that are currently still in a exploratory stage (Figure~\ref{fig:scapplications_extended}).

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\linewidth]{fig/singlecell_technologies_v8.pdf}
	\caption{Developments in computational methods for single-cell omics. Adapted from presentations by Wouter Saelens and myself.}
	\label{fig:scapplications_extended}
\end{figure}

\subsection{Trajectory differential expression}
Trajectory inference methods have made transformative changes in single-cell omics by allowing to study how cells change during a dynamic process of interest in a high-throughput and unsupervised approach. 
A crucial aspect in analysing and interpreting the resulting trajectories is the discovery of genes whose expression significantly shifts in a region of interest within the trajectory, called Trajectory Differential Expression (TDE).

Several trajectory inference methods offer TDE functionality as part of downstream analysis and interpretation of the outputted trajectories \cite{cannoodt_scorpiusimprovestrajectory_2016,qiu_reversedgraphembedding_2017,lonnberg_singlecellrnaseqcomputational_2017,wolf_pagagraphabstraction_2019}. However, current approaches typically cluster cells and perform differential expression analyses between the clusters, or are otherwise limited towards detecting differential expression of a gene along linearly ordered cells along one lineage path in the trajectory.

tradeSeq \cite{vandenberge_trajectorybaseddifferentialexpression_2019} is a generalised TDE method which can be used to discover different types of differential expression along a trajectory, which can be applied as a downstream analysis to trajectories inferred from any TI method. By including a comparative benchmark of TDE methods on real and synthetic data, Van den Berge et al. started a rational dialogue on TDE methodology.

\subsection{Trajectory alignment}
Trajectory alignment allows studying the differences between multiple trajectories that are mostly similar. For example, the cell developmental process of a patient could be compared to that of a healthy control to detect the transcriptomic differences of a particular lineage branch.

Trajectory alignment has been used to compare gene expression kinetics resulting from different biological processes\cite{cacchiarelli_aligningsinglecelldevelopmental_2018}, to compare human, chimpanzee and macaque neuronal development\cite{kanton_organoidsinglecellgenomic_2019}, to find differences in gene regulation in the presence of certain growth factors\cite{mcfaline-figueroa_pooledsinglecellgenetic_2019}, and to compare human and mouse embryogenesis\cite{alpert_alignmentsinglecelltrajectories_2018}.

However, aligning trajectories becomes exponentially more difficult as the complexity of compared trajectories increases. For this reason, trajectory alignment remains a mostly unexplored territory within single-cell omics.

Dynamic Time Warping (DTW) \cite{giorgino_computingvisualizingdynamic_2009} is most commonly used to align linear trajectories. DTW is a technique originating in the field of speech recognition and aligns temporal sequences by creating a warping path between two sequences that indicates which sequence must be dilated or contracted to best match the other one.

Over time, as the affordability of single-cell omics technologies improves and the abundance of available datasets increases, we hypothesize that these methods will become highly relevant in comparing dynamic processes in samples from multiple donors.

\subsection{Differential network inference}
Differential NI methods aim to reconstruct one network for each of the given conditions amongst the transcriptomic profiles. The conditions can be different cellular states or changes in environment, and profiles can be grouped according to prior knowledge or derived through unsupervised clustering. From the resulting networks, one can then investigate the pathways that are differentially activated between conditions (e.g. deregulated pathways between a diseased and healthy condition), or those that are similarly activated between conditions (e.g. similarly activated pathways between two different disease conditions).

Differential NI methods have already been described for bulk -omics data \cite{Ideker2012}, where they have been used, for instance, to elucidate deregulated mechanisms in different subtypes of leukaemia \cite{Gill2014}.

\subsection{Dynamic network inference}
Advances in single-cell transcriptomics have fostered the development of trajectory inference methods \cite{Cannoodt2016}, which are computational tools to reconstruct dynamic processes such as cellular differentiation and cell cycle (Figure \ref{fig:ti_with_ni}). These methods usually recover the underlying structure of the data by focusing on the similarities between cells. The cells can then be ordered along a trajectory that optimally models the data structure.  
This bears resemblance to the more traditional time-series experiments with bulk transcriptomics data, albeit at a much higher temporal resolution. Unlike time-series experiments, trajectories are ideally inferred in an unsupervised manner, as not to introduce human bias.

As in time-series experiments, NI methods can exploit the information provided by trajectory inference methods in order to improve the resulting regulatory network. Several NI methods use this information in order to determine the directionality and the type of an interaction. This has helped to construct more accurate boolean networks (SCNS \cite{Moignard2015}), by selecting the network that most reliably describes the observed cell ordering. Several NI methods use this cell-ordering information to construct ordinary differential equations that are again improved to optimally describe the observed data \cite{Ocone2015,Wei2016}. While such methods are generally able to improve the accuracy of the produced networks by using information from the trajectory, they suffer from the same limitations as discussed at the start of the section since they produce a single regulatory network that is supposed to be a model for all of the cells in the population.

Instead, dynamic NI methods produce a network of interactions with variable activation levels across the trajectory. They provide useful information on which transcription factors are expressed at the beginning of a developmental process, and which gene interactions occur at later stages for instance. Dynamic networks are more complex models in comparison to differential and static networks.

\subsection{Casewise network inference}
Casewise network inference (CNI)
methods\footnote{Other terms are commonly used when dealing with data from a particular source. For example, single-cell NI when applied to single-cell transcriptomics data; sample-specific NI when applied to bulk transcriptomics; patient-derived NI when applied to bulk profiles of patients.}
predict not only which transcription factors regulate which target genes
(Figure~\ref{fig:scgrn}A, top left), but also how active each
interaction is in every case (Figure~\ref{fig:scgrn}A). In CNI, a `case'
might be a cell, but it might also refer to a bulk sample.

While a few pioneering CNI approaches have already been
developed\autocites{aibar_scenicsinglecellregulatory_2017}{kuijjer_estimatingsamplespecificregulatory_2019}{liu_personalizedcharacterizationdiseases_2016},
a quantitative assessment of the performance is hitherto lacking. This
is not surprising, as neither real nor in silico datasets of
cell-specific or even cell-type-specific interactions exists that is
large enough so that it can be used as a ground-truth for evaluating CNI
methods.

\subsection{Intercellular network inference}
Cells are influenced by extracellular signals produced by cells
in their microenvironment. One way of studying intercellular
communication is to profile gene expression in interacting cells.
Especially single-cell-omics technologies are promising because
of their ability to analyze tissue composition in high resolution \cite{yosef_writlargegenomic_2016,tanay_scalingsinglecellgenomics_2017} .
Furthermore, spatial and sophisticated single-cell transcriptomics
technologies have recently been developed to assay the spatial con-
text of a cell \cite{stahl_visualizationanalysisgene_2016,rodriques_slideseqscalabletechnology_2019,eng_transcriptomescalesuperresolvedimaging_2019,vickovic_highdefinitionspatialtranscriptomics_2019} , characterize cellular niches 7 or construct physical
cell–cell interaction maps 8 .
Several studies have recently demonstrated the potential of
single-cell transcriptomics to investigate intercellular communica-
tion 9–14 . There, computational methods, such as CellPhoneDB 14 ,
were applied to infer links between ligands (that is, extracellular
protein signals) expressed by sender cells and receptors expressed
by receiver cells. Thus, these computational approaches elucidate
by which extracellular signals cells can communicate. Nevertheless,
a functional understanding of cell–cell communication requires
knowledge about the influence of these ligand–receptor interac-
tions on target gene expression. Hence, there is a need for compu-
tational methodologies that use expression data of interacting cells
to infer the effects of sender-cell ligands on receiver-cell expression.
Although a few basic methods were developed for bulk data (for
example, CCCExplorer) 15,16 , comprehensive and predictive methods
that are generally applicable to both bulk and single-cell expression
data are lacking.

% When does a cluster represent a new cell type? \cite{kiselev_challengesunsupervisedclustering_2019}
